---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r initialize, echo=FALSE}
rm(list = ls(all.names = T))  # clear the environment
library(lattice)  # load lattice library. required for last question
```


## Loading and preprocessing the data
```{r, echo=TRUE}

if (!file.exists("activity.csv"))
    unzip("activity.zip")


activity <- read.csv(file = "activity.csv", header = TRUE, stringsAsFactors = FALSE, na.strings = "NA")
activity$date <- as.Date(activity$date)

```

## What is mean total number of steps taken per day?
```{r, echo=TRUE}
totalStepsPerDay <- tapply(activity$steps, activity$date, sum, na.rm = TRUE)
hist(totalStepsPerDay, main = "Histogram of Total Steps per Day")
```
  
Mean Steps per Day:
```{r, echo=TRUE}
meanStepsPerDay <- tapply(activity$steps, activity$date, mean, na.rm = TRUE)
meanStepsPerDay
```
  
Median Steps per Day:
```{r, echo=TRUE}
medianStepsPerDay <- tapply(activity$steps, activity$date, median, na.rm = TRUE)
medianStepsPerDay

```

## What is the average daily activity pattern?
```{r, echo=TRUE}

averageNumberOfStepsAcrossDays <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
intervals <- unique(activity$interval)
plot(x = intervals, y = averageNumberOfStepsAcrossDays, type = "l")
intervalIndex <- which.max(averageNumberOfStepsAcrossDays)
intervalValue <- intervals[intervalIndex]
```

The 5-minute interval with the highest number of steps is: `r intervalValue`.  
This interval has `r averageNumberOfStepsAcrossDays[intervalIndex]` steps on average.  

## Imputing missing values

```{r, echo=TRUE}
numRowsWithNAs <- sum(is.na(activity$steps) | is.na(activity$date) | is.na(activity$interval))
```
Number of rows with NAs is: `r numRowsWithNAs`.  
  
The strategy for filling in all of the missing values in the dataset. Using the mean for the equivalent 5-minute interval.
```{r, echo=TRUE}
newActivity <- activity

for (i in nrow(newActivity)) {
  if (is.na(newActivity$steps[i])) {  # Has missing value
    idx <- which(as.integer(row.names(averageNumberOfStepsAcrossDays)) == newActivity$interval[i])  # Find the mean for the same interval
    newActivity$steps[i] <- averageNumberOfStepsAcrossDays[idx] # Impute
  }
  
}

totalStepsPerDay2 <- tapply(newActivity$steps, newActivity$date, sum, na.rm = TRUE)
hist(totalStepsPerDay2, main = "Histogram of Total Steps per Day (with Imputation)")
```
  
Mean Steps per Day (with Imputation):
```{r, echo=TRUE}
meanStepsPerDay2 <- tapply(newActivity$steps, newActivity$date, mean, na.rm = TRUE)
meanStepsPerDay2
```
  
Median Steps per Day (with Imputation):
```{r, echo=TRUE}
medianStepsPerDay2 <- tapply(newActivity$steps, newActivity$date, median, na.rm = TRUE)
medianStepsPerDay2
```


Do these values differ from the estimates from the first part of the assignment?  
Means Differ? `r all.equal(meanStepsPerDay, meanStepsPerDay2)`  
Medians Differ? `r all.equal(medianStepsPerDay, medianStepsPerDay2)`  

Impact?  
It makes the estimates of total steps per day greater or equal to without imputing `r all(totalStepsPerDay2 - totalStepsPerDay >= 0)` 
In addition the means and medians in the for sets with imputation has less number of NAs  

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}
newActivity$weekend <- weekdays(newActivity$date) == "Saturday" | weekdays(newActivity$date) == "Sunday"
newActivity$weekend <- factor(newActivity$weekend, levels = c(FALSE, TRUE), labels = c("weekday", "weekend"))

newActivitySplit <- split(newActivity, newActivity$weekend)

weekdaysAverages <- tapply(newActivitySplit$weekday$steps, newActivitySplit$weekday$interval, mean, na.rm = T)
weekdaysAveragesDS <- data.frame(interval = as.integer(row.names(weekdaysAverages)),
                               steps = weekdaysAverages,
                               weekend = F)

weekendAverages <- tapply(newActivitySplit$weekend$steps, newActivitySplit$weekend$interval, mean, na.rm = T)
weekendAveragesDS <- data.frame(interval = as.integer(row.names(weekendAverages)),
                               steps = weekendAverages,
                               weekend = T)

# recombine averages into a single dataset
fullWeekAverages <- rbind(weekdaysAveragesDS, weekendAveragesDS)
fullWeekAverages$weekend <- factor(fullWeekAverages$weekend, levels = c(FALSE, TRUE), labels = c("weekday", "weekend"))

attach(fullWeekAverages)
xyplot(steps ~ interval | weekend, type = "l", layout=c(1, 2))
```

From the above chart we can easily notice that there are differences between the average steps during weekdays and weekends